---
title: "Class 15: Pertussis miniproject"
author: "PID: A69034741"
format: pdf
editor: visual
---

Pertussis, a.k.a. Whooping Cough, is a highly contagious lung infection caused by the *B. Pertussis* bacteria.

The CDC tracks Pertussis case numbers and they can be accessed [here](https://www.cdc.gov/pertussis/php/surveillance/pertussis-cases-by-year.html?CDC_AAref_Val=https://www.cdc.gov/pertussis/surv-reporting/cases-by-year.html):

We ned to "scrape" this data so we can do stuff with it in R. Let's try the **datapasta** package to do this.

```{r}
# install.packages("datapasta")
# installs an addin
# copy table from URL - Addins -> Paste as data.frame

cdc <- data.frame(
                                 year = c(1922L,1923L,1924L,1925L,
                                          1926L,1927L,1928L,1929L,1930L,1931L,
                                          1932L,1933L,1934L,1935L,1936L,
                                          1937L,1938L,1939L,1940L,1941L,1942L,
                                          1943L,1944L,1945L,1946L,1947L,
                                          1948L,1949L,1950L,1951L,1952L,
                                          1953L,1954L,1955L,1956L,1957L,1958L,
                                          1959L,1960L,1961L,1962L,1963L,
                                          1964L,1965L,1966L,1967L,1968L,1969L,
                                          1970L,1971L,1972L,1973L,1974L,
                                          1975L,1976L,1977L,1978L,1979L,1980L,
                                          1981L,1982L,1983L,1984L,1985L,
                                          1986L,1987L,1988L,1989L,1990L,
                                          1991L,1992L,1993L,1994L,1995L,1996L,
                                          1997L,1998L,1999L,2000L,2001L,
                                          2002L,2003L,2004L,2005L,2006L,2007L,
                                          2008L,2009L,2010L,2011L,2012L,
                                          2013L,2014L,2015L,2016L,2017L,2018L,
                                          2019L,2020L,2021L,2022L, 2024L),
                                 cases = c(107473,164191,165418,152003,
                                          202210,181411,161799,197371,
                                          166914,172559,215343,179135,265269,
                                          180518,147237,214652,227319,103188,
                                          183866,222202,191383,191890,109873,
                                          133792,109860,156517,74715,69479,
                                          120718,68687,45030,37129,60886,
                                          62786,31732,28295,32148,40005,
                                          14809,11468,17749,17135,13005,6799,
                                          7717,9718,4810,3285,4249,3036,
                                          3287,1759,2402,1738,1010,2177,2063,
                                          1623,1730,1248,1895,2463,2276,
                                          3589,4195,2823,3450,4157,4570,
                                          2719,4083,6586,4617,5137,7796,6564,
                                          7405,7298,7867,7580,9771,11647,
                                          25827,25616,15632,10454,13278,
                                          16858,27550,18719,48277,28639,32971,
                                          20762,17972,18975,15609,18617,
                                          6124,2116,3044, 23544)
       )
```

```{r}
cdc
```

```{r}
# install.packages("styler")


```

Let's plot year vs cases to see the trend over time in the US.

```{r}
library(ggplot2)

baseplot <- ggplot(data=cdc) +
  aes (x=year, y=cases) +
  geom_point() +
  geom_line()

```

Let's add the date of wP vaccine rollout across the US

```{r}
baseplot +
  theme_bw() +
  geom_vline(xintercept=1946, col="blue") +
  geom_vline(xintercept=1996, col="red") +
  geom_vline(xintercept=2020, col="gray") +
  geom_vline(xintercept=2004, col="black")

```

## CMI-PB (Computational Models of Immunity - Pertussis Boost)

This project collects and makes freely available doata about the immune response to Pertussis vaccination.

You can access the data via an API which returns JSON format (key:value pairs).

We can use the **jsonlite** package and it's `read_json()` function.

```{r}
# API - application programming interface - way to access data
# install.packages("jsonlite")

library(jsonlite)

subject <- read_json("https://www.cmi-pb.org/api/v5/subject", simplifyVector=TRUE)
```

Let's have a wee peak ad explore this

```{r}
head(subject)
```

>Q. How many subjects do we have?

```{r}
table(subject$biological_sex)
```

> Q. How many wP and aP do we have?

```{r}
table(subject$infancy_vac)
```

> Q. Break down of biological sex and race?

```{r}
table(subject$race, subject$biological_sex)
```

> Q. Does this break down reflect the US population?

```{r}
table(subject$dataset)
```


```{r}
specimen <- read_json("http://cmi-pb.org/api/v5/specimen", simplifyVector = TRUE)
```
```{r}
head(specimen)
```


```{r}
ab_titer <- read_json("http://cmi-pb.org/api/v5/plasma_ab_titer", simplifyVector = TRUE)
```

```{r}
head(ab_titer)
```

We want to merge or "join" these tables so we can have all the info we need about a given antibody measurement.

```{r}
library(dplyr)

meta <- inner_join(subject, specimen)
```

```{r}
head(meta)
```

```{r}
abdata <- inner_join(ab_titer, meta)
```

```{r}
head(abdata)
```

```{r}
nrow(abdata)
```

```{r}
table(abdata$isotype)
```

```{r}
table(abdata$antigen)
```

Let's begin with IgG

```{r}
igg <- filter(abdata, isotype=="IgG")
head(igg)
```
Make a boxplot of IgG antigen levels - this will be a plot of MFI vs antigen

```{r}
library(ggplot2)

ggplot(data=igg) +
  aes (log(MFI_normalised), antigen, col=infancy_vac) +
  geom_boxplot()

```
```{r}
ggplot(data=igg) +
  aes (MFI_normalised, antigen) +
  geom_boxplot() +
  facet_wrap(~infancy_vac)
```

```{r}
table(abdata$visit)
```

```{r}
ggplot(data=igg) +
  aes (MFI_normalised, antigen, col=infancy_vac) +
  geom_boxplot() +
  facet_wrap(~visit)
```


```{r}
filter(igg, antigen=="PT", dataset=="2021_dataset") %>%
ggplot() +
  aes (planned_day_relative_to_boost, MFI_normalised, col=infancy_vac, group=subject_id) +
  geom_point() +
  geom_line() +
  #geom_smooth(aes(group=infancy_vac), method = loess, se = TRUE) +
  geom_vline(xintercept=0, linetype="dashed") +
  geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2021 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)")
  
```

```{r}
filter(igg, antigen=="PT", dataset=="2023_dataset") %>%
ggplot() +
  aes (planned_day_relative_to_boost, MFI_normalised, col=infancy_vac, group=subject_id) +
  geom_point() +
  geom_line() +
 # geom_smooth() +
  geom_vline(xintercept=0, linetype="dashed") +
  geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2021 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)")
```

