---
title: "Class 9: Halloween Candy Mini-Project"
author: "Sawyer Randles, PID: A69034741"
format: pdf
editor: visual
---



```{r}
candy_file <- "candy-data.csv"

candy = read.csv(candy_file, row.names=1)
head(candy)

```

```{r}
nrow(candy)
sum(candy$fruity)
```

> Q1. How many different candy types are in this dataset?

85

> Q2. How many fruity candy types are in the dataset?

38

```{r}
candy["100 Grand", ]$winpercent

```

> Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

100 Grand - 66.97173

> Q4. What is the winpercent value for “Kit Kat”?

76.7686

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

49.6535

```{r}
#install.packages("skimr")
library("skimr")
skim(candy)
```

```{r}
library(dplyr)

candy |>
  filter(rownames(candy) %in% c("Dum Dums", "Twix")) |>
  select(winpercent)
```

> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

winpercent - mean is 2 orders of magnitude higher

> Q7. What do you think a zero and one represent for the candy\$chocolate column?

Zero means it does not contain chocolate, one means it contains chocolate

> Q8. Plot a histogram of winpercent values

```{r}
hist(candy$winpercent, breaks = 10)
```

```{r}
library(ggplot2)

ggplot(candy) +
  aes(winpercent) +
  geom_histogram(bins=10) +
  theme_bw()
```

> Q9. Is the distribution of winpercent values symmetrical?

No - there is a longer tail toward the higher win percent values

> Q10. Is the center of the distribution above or below 50%?

Below

```{r}
candy |>
  filter(winpercent > 75) |>
  filter(pricepercent < 0.5)
```

```{r}
inds <- candy$chocolate == 1
choc.win <- candy[inds,]$winpercent

inds <- candy$fruity == 1
fruit.win <- candy[inds,]$winpercent

```

```{r}
mean(candy$winpercent[as.logical(candy$chocolate)])
mean(candy$winpercent[as.logical(candy$fruity)])
```

```{r}
summary(candy$winpercent)
summary(choc.win)
summary(fruit.win)
```

```{r}
t.test(x= candy$winpercent[as.logical(candy$chocolate)],
       y=(candy$winpercent[as.logical(candy$fruity)])
)
```

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

Chocolate candy is ranked higher than fruit candy on average

> Q12. Is this difference statistically significant?

Yes, with an alpha 0.5

```{r}
candy %>%
  arrange(winpercent) %>%
  tail(5)
```

There are two related fnctions that are useful here `sort()` and `order()`

```{r}
play <- c(2,1,5,3)
sort(play)
```

```{r}
order(play)
```

```{r}
inds <- order(candy$winpercent, decreasing=F)
head(candy[inds,])
```

> Q13. What are the five least liked candy types in this set?

Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, Jawbusters

> Q14. What are the top 5 all time favorite candy types out of this set?

Snickers, Kit Kat, Twix, Reese's Miniatures, Reese's Peanut Butter Cup

> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_col()
```

> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col()
```

> Q. Color your favorite candy your favourite color 

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
my_cols[rownames(candy) == "100 Grand"] = "yellow"

```

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
?rep()
```

> Q17. What is the worst ranked chocolate candy?

Sixlets

> Q18. What is the best ranked fruity candy?

Starburst

```{r}
library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```
> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Reese's Miniatures

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```


> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

Nik L Nip - least popular
Nestle Smarties	
Ring pop		
Hershey's Krackel
Hershey's Milk Chocolate


```{r}
cij <- cor(candy)

#install.packages("corrplot")
library(corrplot)
corrplot(cij)
```

## Principal Component Analysis

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```
```{r}
pca
```


```{r}
plot(pca$x[,1], pca$x[,2], col=my_cols, pch=16)
```

A ggplot version of our PCA plot
```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])
```

```{r}
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```


```{r}
pca$rotation[,1]
```
```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

**NOTES**

The `%in%` operator is useful for checking the intersection of two vectors.

```{r}
c("barry", "liz", "chandra") %in% c("paul", "alice", "liz")
```

```{r}
candy_chocolate <-
  candy |>
  filter(chocolate == 1)

candy_not_chocolate <-
  candy |>
  filter(fruity == 1)
```

```{r}
hist((candy_chocolate$winpercent))
hist((candy_not_chocolate$winpercent))
```

```{r}
play <- c("sawyer", "barry")
rep(play, each = 3)
```
